Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses

### 2021-07-17 manuscript/manuscript.Rmd
Suparna edits

---
title: "Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses"
short-title: "Canary in the Coal Mine"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "April 12, 2021"
published: This is an early draft of a working paper.
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  Many countries across the globe have increasingly cracked down on NGOs over the last few decades. These attempts to repress civil society organizations—many of which are sanctioned by law—represent a bureaucratic form of repression that could indicate more severe human rights abuses and democratic backsliding. This is especially the case for electoral democracies, which unlike autocracies, may not aggressively attack civic space. In this paper, we explore if crackdowns on NGOs predict broader human rights repression. Anti-NGO laws are among the most subtle means of repression and attract the least least domestic and international condemnation, as many audiences mistake this repression as regulation and are not personally affected by it. These laws may also make it easier for states to prevent and deter future large-scale popular mobilization. Using original data on civil society restrictions over the last three decades, we test whether NGO crackdown is a predictor of political terror, physical integrity rights violations, and violations of private civil liberties. We find that while formal de jure laws provide little information in predicting future repression, de facto civil society repression predicts worsening respect for physical integrity rights and civil liberties.
keywords: "civil society restrictions, human rights, political terror, civil liberties"
thanks: >-
  We would like to thank participants of the 2021 International Studies Association (ISA) convention for their comments and feedback.
reference-section-title: References
bibliography: bibliography.bib
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Inter
editor_options: 
  chunk_output_type: console
---

@@

Governments across the globe have been cracking down on civil society organizations (CSOs), a phenomenon commonly referred to the closing or shrinking of civic space [@CarothersBrechenmacher:2014; @DPR:2016; @Chaudhry:2016; @DupuyPrakash:2018]. The last two decades have seen a proliferation of steps designed to limit the influence of CSOs and non-governmental organizations (NGOs), and more than 100 countries have proposed or enacted 244 measures restricting, repressing, or shutting down civil society since 2013 [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory, and over the past decade, the majority of the world has seen a substantial narrowing of allowable civic space. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civil societies [@civicusmonitor; @ChaudhryHeiss:2021].

@@

This hostility to civil society is not unique to autocratic governments or even illiberal democracies [@AbramowitzSchenkkan:2018]—long-established liberal democracies have also increased restrictions on civil society in their countries. Over the past decade, India has canceled the licenses of tens of thousands of NGOs, targeted major groups such as Greenpeace International and Ford Foundation, and has banned 5,000 NGOs from receiving foreign funds [@press-trust-india:2015; @press-trust-india:2019]. In 2013 Canada attempted to revoke the charitable status of civil society organizations for opposing the Northern Gateway pipeline and Canadian mining companies' initiatives abroad [@CIVICUS:2013]. In 2019, activists from the US-based nonprofit No More Deaths were arrested for providing food and water to migrants a the US-Mexico border and faced a 20-year prison sentence before their convictions were overturned in 2020 [@Cramer:2020].

Anti-civil society measures are occasionally violent, but recently, government repression of NGOs has generally been non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of laws to create barriers to NGO advocacy, entry, and funding [@Chaudhry:2016]. The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While in some cases this has led to the depoliticization of NGOs and avoidance of direct political activism by CSOs, states cannot soften or neutralize all forms of political activities in subtle, legal ways like they have with NGOs. Sooner or later, states need to deal with threats to their stability in a more direct, violent, confrontational manner. Administrative NGO laws and general civil society repression could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression in a state, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about the state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as an indicator for subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting human rights abuses, political terror, physical violence, and violations of private civil liberties. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. In some cases, such crackdown can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of civil society repression: official de jure anti-NGO restrictions that are codified in law, and general de facto levels of civil society repression. We propose and test two empirical expectations—that de jure and de factor restrictions predict future political terror, physical integrity rights violations, and government encroachment on private civil liberties.

Using multilevel Bayesian modeling, we find that formal de jure laws provide very little additional information when attempting to predict future repression. However, we find that accounting for more general civil society repression does boost our models' predictive power. Increasing de facto civil society repression predicts an increase in the probability of moderate political terror and results in worsening violations of physical integrity rights and civil liberties.

Our paper makes an important contribution in predicting the onset of repressive state policies. While structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, these factors are also slow to change over time. As such, they may not be particularly suitable for forecasting short-term trajectories, especially for moderately repressive countries. On the other hand, state repression of NGOs is more dynamic and provides information about changing levels of human rights standards and repression in subsequent years.

Below, we first look at existing literature on repression and political violence, outlining state motivations to repress NGOs. We then elaborate on why attacking NGOs may be an important predictor of subsequent repression. Using both original data as well as data from several existing datasets, we build an ensemble of predictive models and explore (1) the marginal effects of civil society repression on different forms of broader human rights repression, and (2) the out-of-sample predictive performance of each model. We conclude with a discussion of our findings and their implications for assessing the repression of human rights.

@@

Repression is the use of coercive action against an individual or organization, within the territorial jurisdiction of the state, for the purpose of imposing a cost on the target as well as deterring specific activities and/or beliefs perceived to be challenging to government personnel, practices, or institutions [@Goldstein:1978, p. xxvii; @PoeTate:1994; @Davenport:2007; and @RitterConrad:2016].

Early literature on repression focused on the role of structural factors such as levels of democracy, economic development, electoral competitiveness in predisposing a state to increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

More recent research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. In addition to attacking groups willingness to organize against states, such preventive repression also damages states' capacity to impose costs [@Davenport:2007a, 47]. NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions or administrative crackdown can also be considered a form of preventive repression [@Chaudhry:2016].

A robust, strong civil society is widely considered as an important component on democratic transition and consolidation [@Putnam:1994]. Politically, as important civil society actors, NGOs enable citizens to voice their grievances, seek redress, and carry out lobbying and advocacy activities to further their interests. They help keep government accountable, monitor human rights violations and corruption. From an economic perspective, NGOs provide services when the state falls short. As a result of these roles, NGOs can wield real influence, which can be threatening to states. State crackdown on NGOs has become pervasive as these organizations have the ability to influence electoral politics [@DPR:2016; @Weiss:2009], aid mobilization [@MurdieBhasin:2011; @BBCM2014; @Boulding:2014], and threaten a state's economic interests [@DietrichMurdie:2017; @LebovicVoeten:2009; @Nielsen:2013].

There is no doubt that violence is often effective at curbing dissent. But it can also backfire, leading to widespread protests, decreasing leaders' legitimacy and increasing their criminal liability [@KimSikkink:2010; @Sikkink:2011]. In contrast to violence, administrative crackdown has fewer adverse consequences—citizens may see it as regulation rather than repression, resulting in little to no counter-mobilization. It may also circumvent concerns among state security or bureaucratic agents about illegitimate orders based on violent repression [@StephanChenoweth:2008]. There are also fewer international consequences since laws—unlike violence—rarely invite condemnation or threats of aid withdrawal [@CarothersBrechenmacher:2014].

We argue that crackdown on NGOs foreshadows future government transgressions. Autocratic leaders want to avoid accountability, and NGOs often stand in conflict with that goal. Civil society organizations, including NGOs, can help shape the narrative of leaders' performance both domestically and internationally. Monitoring by NGOs and subsequent shaming can include significant external revenue loss to states in the form of loss of aid, trade, and foreign direct investments [@KeckSikkink:1998; @BarryClayFlynn:2013]. Government attempts to repress NGOs can signal the beginning of a deteriorating human rights situation. Such repression reflects the government's willingness and capability to use measures that can make it avoid accountability. Once a state targets NGOs, over time it may grow less hesitant to apply repression more widely and more severely.

In exploring the impact of state crackdown on NGOs on subsequent human rights conditions, we look at both *de jure* measures—or formal legislation that seeks to repress NGOs—as well as the *de facto* implementation of these measures. This distinction is important. While states may pass anti-NGO laws, these laws can also lie dormant. For instance, Russia passed its 2015 Undesirable Organizations Law with the purpose of targeting and eliminating specific foreign-connected NGOs it deemed dangerous. One of the law's sponsors, Aleksandr Tarnavsky, described the law as a preventative measure that would not affect the majority of NGOs working in Russia. Rather, the law would be a "weapon hanging on the wall that never fires" and stand as a warning to potentially uncooperative NGOs [@Kozenko:2015].

@@

To measure de jure repression, we use data on anti-NGO laws from @Chaudhry:2016 which includes official NGO laws for 162 countries between 1990 and 2013. We create indexes of the total number of laws passed in each country based on three broad categories of laws related to barriers to entry, funding, and advocacy, and we also sum each of these indexes to create a count of total legal barriers. We also create a yearly indicator variable to show if a country increased its count of anti-NGO laws (models using this outcome are included in the online appendix).

@@

As mentioned previously, simply counting the number of NGO laws does not necessarily capture the restrictiveness of the general civil society environment in a country. Governments might repress civil society without formal laws or pass restrictive laws that they can keep in reserve for later emergencies. To measure the general level of civil society repression in a country, we use V-Dem's CSO repression indicator (`v2csreprss` in @vdem-v10), which captures the extent to which governments harass, deter, liquidate, and arrest members of civil society organizations. Figure \@ref(fig:barriers) shows each of these types of legal barriers per country over time, as well as average levels of civil society repression by regime type. While anti-NGO laws have increased steadily since 1990—particularly laws related to foreign funding—there has not necessarily been a corresponding rise in civil society repression, which has remained fairly constant across different regime types. This divergence is likely indicative of the split between de jure laws and their de facto implementation—a rise in laws is not necessarily tied to direct CSO repression.

@@

To see how both de jure and de facto civil society restrictions influence general human rights repression, we use four different quantitative measures of repression, each of which capture different dimensions of political and civil violence. First, following @GohdesCarey:2017, who use the killing of journalists to predict worsening human rights conditions, we use the Political Terror Scale (PTS), which assigns countries to five different categories of repression each year [@GibneyCornettWood:2019], with Level 5 representing widespread political violence, including arrests, political murders, disappearances, and torture. As seen in the left panel of Figure \@ref(fig:repression), the number of countries at Level 5 has decreased since the 1990s, while most countries are coded as Levels 2 and 3.

@@

Our main question in this paper is whether civil society laws and restrictions serve as an early warning signal to other future forms of repression. As such, we are primarily interested in *predicting* repression and not isolating the causal effect of civil society repression on general human rights abuses. We do not include a complete set of covariates to remove confounding relationships, but instead use a more parsimonious set of explanatory variables that strongly predict repression. Following @GohdesCarey:2017, we control for a country's previous level of repression, its level of democracy as measured by V-Dem's polyarchy index, its logged GDP per capita (measured by the UN), the percent of GDP attributed to trade (measured by the UN), and an indicator for whether the country witnessed more than 25 battle-related deaths (measured by @ucdp2020; @GleditschWallensteenEriksson:2002). Because changes in the human rights environment take time to be reflected by these larger indexes, we model the effect of lagged explanatory variables on repression one year in the future. We also include the lag of civil society laws or repression to account for serial correlation.

We extend @GohdesCarey:2017's previous modeling approach by accounting for country-level variation in repression with multilevel modeling and we include random effects for each country. Accounting for the country this way results in intraclass correlation coefficients of around 0.7 for each of our three outcome variables, which means that the country-based structure of the data explains more than 70% of the variation in outcome, resulting in model estimates that are arguably more precisely measured. Additionally, as a robustness check, we account for within-country and between-country variation by splitting each numeric explanatory variable into its country-level mean and its variation from the mean, following @BellJones:2015. The results from these random effects—within/between (REWB) models are nearly identical to those with the untransformed covariates, so for the sake of simplicity here, we only include results from basic multilevel random effects models.

@@

- **Expectation 1~a~**: Anti-NGO laws (total, advocacy, entry, funding separately) signal worsening political terror
- **Expectation 1~b~**: Anti-NGO laws signal worsening physical violence
- **Expectation 1~c~**: Anti-NGO laws signal constricting private civil liberties
- **Expectation 1~d~**: Anti-NGO laws signal decreasing latent human rights
- **Expectation 2~a~**: Civil society repression signals worsening political terror
- **Expectation 2~b~**: Civil society repression signals worsening physical violence
- **Expectation 2~c~**: Civil society repression signals constricting private civil liberties
- **Expectation 2~d~**: Civil society repression signals decreasing latent human rights

We generate our predictions with two families of Bayesian regression models. Because the Political Terror Scale is measured as an ordered category, we use ordered logistic regression to predict future values of the PTS. As V-Dem's physical violence index and private civil liberties index and latent human rights values are all measured continuously, we predict these values with Gaussian models. We use median values from the models' posterior distributions as our point estimates and provide credible intervals using the 95% highest posterior density.[^bayes] We declare an effect statistically "significant" if the posterior probability of being different from zero exceeds 0.95.

@@

Complete results from all the models (including REWB models), along with posterior predictive checks, goodness-of-fit measures, and prediction diagnostics are all available at a companion website at <https://stats.andrewheiss.com/canary-ngos/>. Our general modeling approach can be summarized as follows:

@@

We present the results of our models in two stages. First, in order to see how minor shifts in civil society laws and repression influence future human rights, we examine the marginal effects of our main explanatory variables when all other model covariates are held constant. These effects are by no means causal—rather, they demonstrate how much variation in predicted human rights is possible as the legal environment for civil society shifts up and down within a typical country. Second, we present predictive diagnostics for each of our models to examine the improvements in predictions. The results of the ordered logistic regression models can be unwieldy to interpret with plain numbers due to varying intercepts and thresholds between PTS levels. Accordingly, we present graphical results for all our models wherever possible and include complete tables of model results in the appendix (see Tables \@ref(tab:tbl-models-pts), \@ref(tab:tbl-models-clphy), \@ref(tab:tbl-models-clpriv), and \@ref(tab:tbl-models-lhr)). For ease in comparing the count of NGO laws with civil society repression, we reverse the x-axis in the marginal effects plots for our second empirical expectation so that increasing the civil society repression represents *worse* repression rather than less.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of NGO laws results in this turn from Level 1 to Levels 2 or 3, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of legal barriers, the overall probability of seeing Levels 4 and 5 is minuscule, and the adding new laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5. Specifically, in addition to measuring the scope and intensity of government violence, the PTS scale also looks at the proportion of population targeted for abuse. Countries that receive a 5 are rare, as this score captures physical abuses routinely perpetrated against citizens not involved in politics, and includes cases where governments engage in "scorched earth" counterinsurgency policies or summary executions based on class or ethnic affiliation [@GibneyCornettWood:2019]. The majority of movement across PTS categories should thus occur between Levels 1–3.

>>>>

---
title: "Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses"
short-title: "Canary in the Coal Mine"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "April 12, 2021"
published: This is an early draft of a working paper.
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  An increasing number of countries cracked down on NGOs over the last few decades. These attempts to repress civil society organizations—many of which are sanctioned by law—represent a bureaucratic form of repression that could indicate more severe human rights abuses and democratic backsliding in the future. This is especially the case for electoral democracies, which unlike autocracies, may not aggressively attack civic space. In this paper, we explore if crackdowns on NGOs predict broader human rights repression. Anti-NGO laws are among the most subtle means of repression and attract much less domestic and international condemnation compared to the use of violence, as many audiences mistake this repression as regulation. These laws may also make it easier for states to prevent future large-scale popular mobilization. Using original data on civil society restrictions over the last three decades, we test whether NGO crackdown is a predictor of political terror, physical integrity rights violations, and violations of private civil liberties. We find that while formal de jure laws provide little information in predicting future repression, their patterns of implementation or de facto civil society repression predicts worsening respect for physical integrity rights and civil liberties.
keywords: "civil society, NGOs, repression, human rights, political terror, civil liberties"
thanks: >-
  We would like to thank participants of the 2021 International Studies Association (ISA) convention, and the Human Rights on the Edge: The Future of International Human Rights Law and Practice hosted by the Global Human Rights Hub at Arizona State University for their comments and feedback.
reference-section-title: References
bibliography: bibliography.bib
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Inter
editor_options: 
  chunk_output_type: console
---

@@

Over the last few decades, governments across the globe have been increasinly cracking down on civil society organizations (CSOs), particularly non-governmental organizations (NGOs) -- a phenomenon commonly referred to the closing or shrinking of civic space [@CarothersBrechenmacher:2014; @DPR:2016; @Chaudhry:2016; @DupuyPrakash:2018]. More than 100 countries have proposed or enacted 244 measures restricting, repressing, or shutting down civil society since 2013 [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory, and over the past decade, the majority of the world has seen a substantial narrowing of allowable civic space. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civil societies [@civicusmonitor; @ChaudhryHeiss:2021].[^open]

@@

This hostility to civil society is not unique to autocratic governments or even illiberal democracies [@AbramowitzSchenkkan:2018]—long-established liberal democracies have also increased restrictions on civil society in their countries. Over the past decade, India has canceled the licenses of tens of thousands of NGOs, targeted major groups such as Greenpeace International, Amnesty International and Ford Foundation, and has banned 5,000 NGOs from receiving foreign funds [@press-trust-india:2015; @press-trust-india:2019]. In 2013 Canada attempted to revoke the charitable status of NGOs for opposing the Northern Gateway pipeline and Canadian mining companies' initiatives abroad [@CIVICUS:2013]. In 2019, activists from the US-based nonprofit No More Deaths were arrested for providing food and water to migrants a the US-Mexico border and faced a 20-year prison sentence before their convictions were overturned in 2020 [@Cramer:2020].

Much of this targeting of NGOs has been non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of laws to create barriers to NGO advocacy, entry, and funding [@Chaudhry:2016]. The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While this can lead to the depoliticization and avoidance of direct political activism by CSOs,[CITE] states cannot soften or neutralize all forms of political activity and advocacy in subtle, legal ways like they have with NGOs. Sooner or later, states may need to deal with threats to their stability in a more direct, violent, confrontational manner. Anti-NGO laws and repression of CSOs could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression in a state, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about the state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as a predictor of subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting human rights abuses, political terror, physical violence, and violations of private civil liberties. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. It can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of NGO repression: official de jure anti-NGO restrictions that are codified in law, and their de facto implementation. Using multilevel Bayesian modeling, we find that formal de jure laws provide very little additional information when attempting to predict future repression. However, we find that accounting for their implementation does boost our models' predictive power. Increasing de facto civil society repression predicts an increase in the probability of moderate political terror and results in worsening violations of physical integrity rights and civil liberties.

Our paper makes an important contribution in predicting the onset of repressive state policies. While structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, these factors are also slow to change over time. As such, they may not be particularly suitable for forecasting short-term trajectories, especially for moderately repressive countries. On the other hand, state repression of NGOs is more dynamic process than may provide insight into changing levels of human rights standards and repression in subsequent years. While recent research has made important contributions to studying the extent and effects of state crackdown on NGOs, much of this focuses on de jure restrictions and its impact on NGOs and donors.[CITE] By looking at both laws, as well as their implementation when combined with extra-legal means, we are able to assess the long-term effects of closing civic space on broader patterns of repression as well.

Below, we first look at existing literature on repression and political violence, outlining state motivations to repress NGOs. We then elaborate on why attacking NGOs may be an important predictor of subsequent repression. Using both original data as well as data from several existing datasets, we build an ensemble of predictive models and explore (1) the marginal effects of civil society repression on different forms of broader human rights repression, and (2) the out-of-sample predictive performance of each model. We conclude by discussing the implications of our findings on the role of NGOs in the international community and prospects of democratic backsliding.

@@

Repression is the use of coercive action against an individual or organization, within the territorial jurisdiction of the state, for the purpose of imposing a cost on the target as well as deterring specific activities and/or beliefs perceived to be challenging to government personnel, practices, or institutions [@Goldstein:1978, p. xxvii; @Davenport:2007, 487; and @RitterConrad:2019, 15]. Early literature on repression focused on the role of structural factors such as levels of democracy, economic development, electoral competitiveness in predisposing a state to increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

More recent research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. In addition to attacking groups willingness to organize against states, such preventive repression also damages states' capacity to impose costs [@Davenport:2007a, 47]. NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions can be considered a form of preventive repression [@Chaudhry:2016].

While many saw the initial proliferation of NGOs as leading to a "symbiotic relationship" between NGOs and states due to their mutual goals, this perception changed over time (Reimann 2006, 63). With expanded powers and their perceived role in the Color Revolutions, NGOs began to be seen as having a real influence. (CITE) State crackdown on NGOs became pervasive as these organizations have the ability to influence electoral politics [@DPR:2016; @Weiss:2009], aid mobilization [@MurdieBhasin:2011; @BBCM2014; @Boulding:2014], and threaten a state's economic interests [@DietrichMurdie:2017; @LebovicVoeten:2009; @Nielsen:2013].

Given this threat, a number of states decided to repress threatening NGOs. While violence is often effective at curbing dissent, it can also backfire, leading to widespread protests, decreasing leaders' legitimacy and increasing their criminal liability [@KimSikkink:2010; @Sikkink:2011]. In contrast to violence, administrative crackdown has fewer adverse consequences—citizens may see it as regulation rather than repression, resulting in little to no counter-mobilization. It may also circumvent concerns among state security or bureaucratic agents about illegitimate orders based on violent repression [@StephanChenoweth:2008]. There are also fewer international consequences since laws—unlike violence—rarely invite condemnation or threats of aid withdrawal [@CarothersBrechenmacher:2014].

This administrative crackdown via anti-NGO laws may foreshadow future government transgressions. Autocratic leaders want to avoid accountability, and NGOs often stand in conflict with that goal. Civil society organizations, including NGOs, can help shape the narrative of leaders' performance both domestically and internationally. Monitoring by NGOs and subsequent shaming can include significant external revenue loss to states in the form of loss of aid, trade, and foreign direct investments [@KeckSikkink:1998; @BarryClayFlynn:2013]. Government attempts to repress NGOs can signal the beginning of a deteriorating human rights situation. Such repression reflects the government's willingness and capability to use measures that can make it avoid accountability. Once a state targets NGOs, over time it may grow less hesitant to apply violent repression more widely and more severely.

In exploring the impact of state crackdown on NGOs on subsequent human rights conditions, we look at both *de jure* measures—or formal legislation that seeks to repress NGOs—as well as the *de facto* implementation of these measures. This distinction is important. While states may pass anti-NGO laws, these laws can also lie dormant. For instance, Russia passed its 2015 Undesirable Organizations Law with the purpose of targeting and eliminating specific foreign-connected NGOs it deemed dangerous. One of the law's sponsors, Aleksandr Tarnavsky, described the law as a preventative measure that would not affect the majority of NGOs working in Russia. Rather, the law would be a "weapon hanging on the wall that never fires" and stand as a warning to potentially uncooperative NGOs [@Kozenko:2015]. Similarly, Egypt's 2002 NGO law instated severe barriers to freedom of association, banned NGOs from engaging in activities that "threaten national unity or violate public order or morality," and also prevented groups for receiving foreign funding. But despite the harsh legal framework, the NGO sector nearly doubled in size between 1993-2011. This was because of the state's limited targeting of groups working on politically sensitive issues such as human rights and democracy, while tolerating other kinds of organizations (@Brechenmacher 2017).

@@

To measure de jure repression, we use data on anti-NGO laws from @Chaudhry:2016 which includes official NGO laws for all countries (excluding micro-states) between 1990 and 2013. We create indeces of the total number of laws passed in each country based on three broad categories of laws related to barriers to entry, funding, and advocacy, and we also sum each of these indeces to create a count of total legal barriers. We also create a yearly indicator variable to show if a country increased its count of anti-NGO laws (models using this outcome are included in the online appendix).

@@

To measure de facto civil society repression in a country, we use Varieties of Democracy's (V-Dem) CSO repression indicator (`v2csreprss` in @vdem-v10), which captures the extent to which governments harass, deter, liquidate, and arrest members of civil society organizations. Figure \@ref(fig:barriers) shows each of these types of legal barriers per country over time, as well as average levels of de facto civil society repression by regime type. While anti-NGO laws have increased steadily since 1990—particularly laws related to foreign funding—there has not necessarily been a corresponding rise in de facto civil society repression, which has remained fairly constant across different regime types. This divergence is likely indicative of the split between de jure laws and their de facto implementation—a rise in laws is not necessarily tied to direct CSO repression.

@@

To see how both de jure and de facto civil society restrictions influence general human rights repression, we use four different quantitative measures of repression, each of which capture different dimensions of political and civil violence. First, following @GohdesCarey:2017, who use the killing of journalists to predict worsening human rights conditions, we use the Political Terror Scale (PTS), which assigns countries to five different categories of repression each year [@GibneyCornettWood:2019], with Level 5 representing widespread political violence, including arrests, political murders, disappearances, and torture. As seen in the left panel of Figure \@ref(fig:repression), the number of countries at Level 5 has decreased since the 1990s, while most countries are coded as Levels 2 and 3 (moderate levels of political terror).

@@

Since we look at whether anti-NGO laws and their implementation can serve as an early warning signal for broader repression, we are primarily interested in *predicting* repression and not isolating the causal effect of civil society repression on general human rights abuses. We do not include a complete set of covariates to remove confounding relationships, but instead use a more parsimonious set of explanatory variables that strongly predict repression. Following @GohdesCarey:2017, we control for a country's previous level of repression, its level of democracy as measured by V-Dem's polyarchy index, its logged GDP per capita (measured by the UN), the percent of GDP attributed to trade (measured by the UN), and an indicator for whether the country witnessed more than 25 battle-related deaths (measured by @ucdp2020; @GleditschWallensteenEriksson:2002). Because changes in the human rights environment take time to be reflected by these larger indexes, we model the effect of lagged explanatory variables on repression one year in the future. We also include the lag of civil society laws or repression to account for serial correlation.

We extend @GohdesCarey:2017's previous modeling approach by accounting for country-level variation in repression with multilevel modeling and we include random effects for each country. Accounting for the country this way results in intraclass correlation coefficients of around 0.7 for each of our three outcome variables, which means that the country-based structure of the data explains more than 70% of the variation in outcome, resulting in model estimates that are arguably more precisely measured. Additionally, as a robustness check, we account for within-country and between-country variation by splitting each numeric explanatory variable into its country-level mean and its variation from the mean, following @BellJones:2015. [@TODO: Footnote this REWB stuff] The results from these random effects—within/between (REWB) models are nearly identical to those with the untransformed covariates, so for the sake of simplicity here, we only include results from basic multilevel random effects models.

@@

- **Expectation 1~a~**: Anti-NGO laws (total, advocacy, entry, funding separately) signal worsening political terror
- **Expectation 1~b~**: Anti-NGO laws signal worsening physical violence
- **Expectation 1~c~**: Anti-NGO laws signal constricting civil liberties
- **Expectation 1~d~**: Anti-NGO laws signal decreasing latent human rights respect
- **Expectation 2~a~**: Civil society repression signals worsening political terror
- **Expectation 2~b~**: Civil society repression signals worsening physical violence
- **Expectation 2~c~**: Civil society repression signals constricting civil liberties
- **Expectation 2~d~**: Civil society repression signals decreasing latent human rights respect

We generate our predictions with two families of Bayesian regression models. Because the Political Terror Scale is measured as an ordered category, we use ordered logistic regression to predict future values of the PTS. As V-Dem's physical violence index and private civil liberties index and latent human rights values are all measured continuously, we predict these values with Gaussian models. We use median values from the models' posterior distributions as our point estimates and provide credible intervals using the 95% highest posterior density.[^bayes] We declare an effect statistically "significant" if the posterior probability of being different from zero (i.e. above or below) exceeds 0.95.

[^open]: Defined as the state both enabling and safeguarding citizens' freedom to form associations, authorities being tolerant of criticism from CSOs and laws governing freedom of peaceful assembly adhere to international laws and standards.

@@

@TODO: FOOTNOTE or APPENDIX: Complete results from all the models (including REWB models), along with posterior predictive checks, goodness-of-fit measures, and prediction diagnostics are all available at a companion website at <https://stats.andrewheiss.com/canary-ngos/>. Our general modeling approach can be summarized as follows:

@@

We present the results of our models in two stages. First, to see how minor shifts in civil society laws and repression influence future human rights, we examine the marginal effects of our main explanatory variables when all other model covariates are held constant. These effects are by no means causal—rather, they demonstrate how much variation in predicted human rights is possible as the legal environment for civil society improves or deteriorates within a typical country. Second, we present predictive diagnostics for each of our models to examine the improvements in predictions. The results of the ordered logistic regression models can be unwieldy to interpret with plain numbers due to varying intercepts and thresholds between PTS levels. Accordingly, we present graphical results for all our models wherever possible and include complete tables of model results in the appendix (see Tables \@ref(tab:tbl-models-pts), \@ref(tab:tbl-models-clphy), \@ref(tab:tbl-models-clpriv), and \@ref(tab:tbl-models-lhr)). For ease in comparing the count of NGO laws with civil society repression, we reverse the x-axis in the marginal effects plots for our second empirical expectation so that increasing the civil society repression represents *worse* repression rather than less.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of NGO laws results in this turn from Level 1 to Levels 2 or 3, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of legal barriers, the overall probability of seeing Levels 4 and 5 is minuscule, and the adding new laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5. Specifically, in addition to measuring the scope and intensity of government violence, the PTS scale also looks at the proportion of population targeted for abuse. Countries that receive a 4 and 5 are rare, as 4 denotes that murders, disappearances, and torture are a common part of life, while 5 denotes that physical abuses routinely perpetrated against citizens not involved in politics, and includes cases where governments engage in "scorched earth" counterinsurgency policies or summary executions based on class or ethnic affiliation [@GibneyCornettWood:2019]. The majority of movement across PTS categories should thus occur between Levels 1–3.
